---
title: "Tulare Basin Inorganic Water Quality Maps"
author: "Rich Pauloo"
date: "July 23, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##### This analysis explores inorganic (TDS) groundwater contamination in the Tulare Basin, California. Data was generously provided by [CV-SALTS](https://www.cvsalinity.org/) and [CDM Smith](https://cdmsmith.com/), and aggregated from:

* United States Geologic Survey
* California Department of Public Health  
* Groundwater Ambient Monitoring and Assessment Program  
* Environmental Defense Fund  
* California Department of Water Resources  
* independent dairy farms monitoring wells  

##### Data ranges from 1902 - 2014. Later years (1980 - 2014) are much more data rich than early years (1902 - 1940).  

##### All data cleaning, analysis, and mapping was carried out within R version 1.0.143.  
  
  
  
## Set working directory, load requried packages, and load data
```{r set wd, load required packages, and load data, warnings = FALSE, messages = FALSE, results="hide"}
setwd("E:/rpauloo/Documents/GitHub/Contaminant-Mapping")

required.packages <- c("sp", "XML", "rgdal", "rgeos", "raster", "ggplot2", 
                       "dismo", "leaflet", "RColorBrewer", "classInt")
suppressMessages(lapply(required.packages, require, character.only = TRUE))

dat = read.csv(file = "TDS_Features_Tulare_Clip_TableToExcel.csv", stringsAsFactors = FALSE)

dat = dat[, c("Well_ID", "IAZ", "Database", "Samp_Date", 
              "Year", "Time_perio", "Result", "Latitude",
              "Longitude", "Depth_ft", "Well_Type", "DepthClass")]
```


## Clean Data
```{r Clean data, warnings = FALSE, messages = FALSE}
library(plyr)
str(dat)

# create logical arrays to define sampling time intervals
time1 = ifelse(dat$Year <= 1940, 1, NA) 
time2 = ifelse(dat$Year > 1940 & dat$Year <= 1980, 1, NA) 
time3 = ifelse(dat$Year > 1980, 1, NA) 

record.length = lapply(list(time1, time2, time3), sum, na.rm = TRUE)
# 160 records from 1902-1940
# 19,894 records from 1940-1980
# 34,117 records from 1980-2014

# substitute logical values from time1, time2, time3 with character vectors for plyr
time1 = as.numeric(time1)
time2 = as.numeric(sub(1, 2, time2))
time3 = as.numeric(sub(1, 3, time3))
tframe = cbind(time1, time2, time3)
tframe = rowSums(tframe[,1:3], na.rm = TRUE)

# ensure all values are in tframe
unique(tframe)
table(tframe)

# add tframe to dat
dat$tframe = tframe

# filter for Shallow and Deep depth class
unique(dat$DepthClass) # need to get rid of unknown values
fildat = subset(dat, DepthClass == "Deep" | DepthClass == "Shallow") # subset deep/shallow
unique(fildat$DepthClass) # shows that Deep and Shallow are the only depth classes now
records.trimmed = nrow(dat) - nrow(fildat) # trimmed 23,209 observations without a depth class

# ddply by well_ID, year, depthclass, and tframe and take Result means 
columns = c("Well_ID", "DepthClass", "tframe", "Latitude", "Longitude")
plydat = ddply(fildat, columns, summarize, Result = round(mean(Result), digits=2))
```


## Data Analysis and creation of spatial points df
```{r create spatial points df, warnings = FALSE, messages = FALSE}
well_coords = cbind(plydat$Longitude, plydat$Latitude)
pts = SpatialPoints(well_coords)
# deinfe color palette to be used. blue values correspond to fresh water, red to salty water
co = c("blue","steelblue2","seashell1","orangered2","red")
```


## Histogram of all results
##### Data is heavily skewed to the right. This skew blurs interesting data in the freshwater 0 - 1,000 mg/L range. Bin sizes in the final map should reflect this distribution of data and not bias the graph towards the outliers. A lognormal plot is insufficient because TDS is typically not reported in log scale.
```{r Histogram of all results, warnings = FALSE, messages = FALSE}
ggplot(plydat, aes(Result)) + 
  geom_histogram(binwidth = 100) + 
  xlab("TDS (mg/L)") +
  ggtitle("Histogram of TDS observations in the Tulare Basin (1902-2014)") +
  theme_bw()
```


## Define Map Function
```{r map function, warnings = FALSE, messages = FALSE}
# filter plydat for only observations in time frame t and depth class d
map.TDS = function(t, d){
t.dc = plydat[, c(1:3,6)]
t.dc = subset(t.dc, DepthClass == d & 
                    tframe == t) # remove lat and long from df, and filter for desired results

# create logical vector for all t and d indexes to extract coordinates of these observations
t.dc.pts = ifelse(plydat$DepthClass == d & 
                  plydat$tframe == t, TRUE, FALSE)

# use logical vector to subset well_coords, and create spatial points data frame for time frame t and depth d
ptsdf = SpatialPointsDataFrame(pts[t.dc.pts , ], data = t.dc)

# define palette based on time frame t and depth class d observations
# set bins with a vector of breaks, so they are fine from 0 - 1,000 and coarse from 1,000 - 50,000
pal = colorBin(palette = co,
         domain = t.dc$Result, bins = c(0,200,400,600,800,1000,
                                        5000,10000,50000))

# overwite t argument inside the function for the legend to display time frame
if(t == 3){t = "(1980-2014)"}
if(t == 2){t = "(1940-1980)"}
if(t == 1){t = "(1902-1940)"}

# generate map assocaited with time frame t and depth d
map = leaflet(data = ptsdf) 
map = addTiles(map)
map %>%
  addCircleMarkers(map,
    lng = well_coords[t.dc.pts,1], # longitude
    lat = well_coords[t.dc.pts,2], # latitude
    radius = 4, # fixed radius size
    color = ~pal(Result),
    stroke = FALSE, 
    fillOpacity = 0.8,
    popup = paste(as.character(ptsdf$Result), 
                  "mg/L TDS", 
                  by = " ")
  ) %>%
  
  addLegend("topright", pal = pal,
            values = ~Result,
            title = paste(d, t, sep = " "), # legend displays time frame and depth class as title
            labFormat = labelFormat(suffix = " mg/L"),
            opacity = 1
  ) %>%
  
  addProviderTiles(providers$Esri.WorldTerrain) # override default map
}
```


```{r Apply Map Function, include = FALSE}
# loop over time frames (1-3) and depths (shallow, deep)
for(i in 1:3){
  for(j in c("Shallow", "Deep")){
    print(list(map.TDS(i,j)))
  }
}

```

## Specify character vectors for plotting depth
```{r, warnings = FALSE, messages = FALSE}
d1 = "Shallow"
d2 = "Deep"
```


## Shallow Aquifer (1902 - 1940)
##### Click on points to view TDS values.
```{r, warnings = FALSE, messages = FALSE}
map.TDS(1, d1)
```


## Deep Aquifer (1902 - 1940)
##### Click on points to view TDS values.
```{r, warnings = FALSE, messages = FALSE}
map.TDS(1, d2)
```


## Shallow Aquifer (1940 - 1980)
##### Click on points to view TDS values.
```{r, warnings = FALSE, messages = FALSE}
map.TDS(2, d1)
```


## Deep Aquifer (1940 - 1980)
##### Click on points to view TDS values.
```{r, warnings = FALSE, messages = FALSE}
map.TDS(2, d2)
```


## Shallow Aquifer (1980 - 2014)
##### Click on points to view TDS values.
```{r, warnings = FALSE, messages = FALSE}
map.TDS(3, d1)
```


## Deep Aquifer (1980 - 2014)
##### Click on points to view TDS values.
```{r, warnings = FALSE, messages = FALSE}
map.TDS(3, d2)
```






